---
title: "Melbourne Housing Data Exploration"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
editor: visual
---

## Setup

```{r}
#| message: false
#| warning: false

library(tidyverse)
```

## Import and Clean Data

```{r}
#| message: false

housing <- read_csv(
  "data/Melbourne_housing_FULL.csv",
  na = c("", "NA", "#N/A"),
  show_col_types = FALSE
)

housing <- housing |>
  mutate(
    Type = recode(
      Type,
      "h" = "House",
      "u" = "Unit",
      "t" = "Townhouse"
    )
  )
```

The Melbourne Housing dataset contains information about residential properties in Melbourne, Australia. The variables include property characteristics, geographic information, sales information, and property prices.

## Initial Data Exploration

The dataset contains 34,857 property records and 21 variables.

```{r}
dim(housing)
```

```{r}
names(housing)
```

```{r}
glimpse(housing)
```

## Missing Values

Some variables in the dataset contain a substantial number of missing observations. The variables that will be used most heavily in the application are checked below.

```{r}
housing |>
  summarise(
    missing_price = sum(is.na(Price)),
    missing_rooms = sum(is.na(Rooms)),
    missing_type = sum(is.na(Type)),
    missing_region = sum(is.na(Regionname))
  )
```

`Rooms` and `Type` contain no missing observations, while `Price` has a relatively large number of missing values. Missing price observations will be excluded when price is needed for a summary or visualization.

# Categorical Variable Summaries

## One-Way Table: Property Type

```{r}
table(
  housing$Type,
  useNA = "ifany"
)
```

Houses are the most common property type in the dataset, followed by units and townhouses.

## One-Way Table: Region

```{r}
table(
  housing$Regionname,
  useNA = "ifany"
)
```

The observations are distributed across eight Melbourne regions. Southern Metropolitan and Northern Metropolitan contain the largest numbers of properties.

## Two-Way Table: Property Type and Region

```{r}
table(
  housing$Type,
  housing$Regionname,
  useNA = "ifany"
)
```

The two-way table shows that the distribution of property types differs considerably across regions.

# Numerical Summaries

## Property Price by Property Type

```{r}
housing |>
  filter(
    !is.na(Price),
    !is.na(Type)
  ) |>
  group_by(Type) |>
  summarise(
    count = n(),
    mean_price = mean(Price),
    median_price = median(Price),
    sd_price = sd(Price),
    min_price = min(Price),
    max_price = max(Price)
  )
```

Houses have the highest mean and median prices, while units have the lowest typical prices.

## Property Price by Region

```{r}
housing |>
  filter(
    !is.na(Price),
    !is.na(Regionname)
  ) |>
  group_by(Regionname) |>
  summarise(
    count = n(),
    mean_price = mean(Price),
    median_price = median(Price),
    sd_price = sd(Price)
  ) |>
  arrange(desc(mean_price))
```

Southern Metropolitan has the highest mean and median property prices among the regions in the dataset.

## Number of Rooms by Property Type

```{r}
housing |>
  filter(
    !is.na(Type)
  ) |>
  group_by(Type) |>
  summarise(
    mean_rooms = mean(Rooms),
    median_rooms = median(Rooms),
    sd_rooms = sd(Rooms)
  )
```

Houses tend to contain more rooms than townhouses or units.

# Graphical Exploration

```{r}
#| include: false

price_limit <- quantile(
  housing$Price,
  0.99,
  na.rm = TRUE
)
```

The 99th percentile of price is used as an upper plotting limit for several visualizations. This prevents a small number of extremely expensive properties from compressing the majority of observations.

## Plot 1: Number of Properties by Type

```{r}
ggplot(
  housing |>
    filter(!is.na(Type)),
  aes(x = Type)
) +
  geom_bar() +
  labs(
    title = "Number of Properties by Property Type",
    x = "Property Type",
    y = "Number of Properties"
  ) +
  theme_minimal()
```

Houses represent the majority of properties in the dataset, while townhouses are the least common property type.

## Plot 2: Distribution of Property Prices

```{r}
housing |>
  filter(
    !is.na(Price),
    Price <= price_limit
  ) |>
  ggplot(
    aes(x = Price)
  ) +
  geom_histogram(
    bins = 40
  ) +
  scale_x_continuous(
    labels = scales::dollar
  ) +
  labs(
    title = "Distribution of Melbourne Property Prices",
    subtitle = "Prices above the 99th percentile are excluded for readability",
    x = "Property Price",
    y = "Number of Properties"
  ) +
  theme_minimal()
```

Property prices have a right-skewed distribution, with most properties concentrated below roughly two million dollars and fewer observations at higher prices.

## Plot 3: Property Price by Property Type

```{r}
housing |>
  filter(
    !is.na(Price),
    !is.na(Type),
    Price <= price_limit
  ) |>
  ggplot(
    aes(
      x = Type,
      y = Price,
      fill = Type
    )
  ) +
  geom_boxplot(
    show.legend = FALSE
  ) +
  scale_y_continuous(
    labels = scales::dollar
  ) +
  labs(
    title = "Property Price by Property Type",
    subtitle = "Prices above the 99th percentile are excluded for readability",
    x = "Property Type",
    y = "Property Price"
  ) +
  theme_minimal()
```

Houses generally have higher prices than townhouses and units. Units have the lowest median property price.

## Plot 4: Property Price and Distance from Melbourne CBD

```{r}
housing |>
  filter(
    !is.na(Price),
    !is.na(Distance),
    !is.na(Type),
    Price <= price_limit
  ) |>
  ggplot(
    aes(
      x = Distance,
      y = Price,
      color = Type
    )
  ) +
  geom_point(
    alpha = 0.30
  ) +
  scale_y_continuous(
    labels = scales::dollar
  ) +
  labs(
    title = "Property Price and Distance from Melbourne CBD",
    subtitle = "Points are colored by property type",
    x = "Distance from Melbourne CBD",
    y = "Property Price",
    color = "Property Type"
  ) +
  theme_minimal()
```

Higher-priced properties appear more frequently closer to Melbourne's central business district. Property type also helps explain some of the variation in price.

## Plot 5: Median Price by Region and Property Type

```{r}
region_type_price <- housing |>
  filter(
    !is.na(Price),
    !is.na(Regionname),
    !is.na(Type)
  ) |>
  group_by(
    Regionname,
    Type
  ) |>
  summarise(
    median_price = median(Price),
    .groups = "drop"
  )

ggplot(
  region_type_price,
  aes(
    x = reorder(
      Regionname,
      median_price
    ),
    y = median_price,
    fill = Type
  )
) +
  geom_col(
    position = "dodge"
  ) +
  coord_flip() +
  scale_y_continuous(
    labels = scales::dollar
  ) +
  labs(
    title = "Median Property Price by Region and Property Type",
    x = "Region",
    y = "Median Property Price",
    fill = "Property Type"
  ) +
  theme_minimal()
```

Median prices vary across both geographic region and property type. Southern Metropolitan and Eastern Metropolitan tend to contain the most expensive properties.

## Plot 6: Property Price and Number of Rooms by Region

```{r}
housing |>
  filter(
    !is.na(Price),
    !is.na(Rooms),
    !is.na(Regionname),
    Price <= price_limit
  ) |>
  ggplot(
    aes(
      x = Rooms,
      y = Price
    )
  ) +
  geom_point(
    alpha = 0.25
  ) +
  facet_wrap(
    ~ Regionname
  ) +
  scale_y_continuous(
    labels = scales::dollar
  ) +
  labs(
    title = "Property Price and Number of Rooms by Region",
    subtitle = "Each panel represents a Melbourne region",
    x = "Number of Rooms",
    y = "Property Price"
  ) +
  theme_minimal()
```

The relationship between the number of rooms and property price differs across Melbourne regions. The faceted display makes these regional patterns easier to compare.

## Plot 7: Heatmap of Median Property Prices

```{r}
heatmap_data <- housing |>
  filter(
    !is.na(Price),
    !is.na(Regionname),
    !is.na(Type)
  ) |>
  group_by(
    Regionname,
    Type
  ) |>
  summarise(
    median_price = median(Price),
    .groups = "drop"
  )

ggplot(
  heatmap_data,
  aes(
    x = Type,
    y = Regionname,
    fill = median_price
  )
) +
  geom_tile() +
  scale_fill_viridis_c(
    labels = scales::dollar
  ) +
  labs(
    title = "Median Property Prices by Region and Property Type",
    x = "Property Type",
    y = "Region",
    fill = "Median Price"
  ) +
  theme_minimal()
```

The heatmap provides a compact comparison of median property prices across combinations of region and property type. The highest median prices are concentrated among houses in the more expensive metropolitan regions.

# Variables Selected for the Shiny Application

Based on the initial exploration, the following variables will work well for the main filtering controls in the Shiny application:

-   **Property Type (`Type`)** — categorical
-   **Region (`Regionname`)** — categorical
-   **Price (`Price`)** — quantitative
-   **Number of Rooms (`Rooms`)** — quantitative

Property type and region provide meaningful categories for users to explore, while price and number of rooms are intuitive numerical characteristics of residential properties.

# Summary

The Melbourne Housing dataset contains substantial variation in property prices, geographic location, property type, and housing characteristics.

The exploratory analysis suggests that both **region** and **property type** are strongly associated with differences in property prices. Properties closer to Melbourne's central business district also tend to include a greater concentration of higher-priced homes.

These patterns make the dataset well suited for an interactive Shiny application where users can subset properties and dynamically explore categorical summaries, numerical summaries, and visualizations.
